A GPU Sonar Simulator for Automatic Target Recognition

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:04:30	12:09:07	(Wed30.Apr)

\title{A GPU Sonar Simulator for Automatic Target Recognition}%
%
\author{Jo Inge Buskenes\firstAddress, \O{}ivind Midtgaard\secondAddress, Herman Midelfart\secondAddress}%
%
\begin{contact}
  \firstAddress Dept. of Informatics, P.O.Box 1080 Blindern, N-0316 OSLO\\
  \secondAddress Norwegian Defence Research Establishment (FFI) P.O.Box 25, NO-2027 Kjeller
\end{contact}%
%
\begin{contact}
Contact author: Jo Inge Buskenes\\
Dept. of Informatics, Univ. of Oslo, P.Box 1080 Blindern, N-0316 OSLO\\
\href{mailto:joibu@ifi.uio.no}{joibu@ifi.uio.no}
\end{contact}%
%
\begin{abstract}
Template matching is a common technique for automatic classification of objects in synthetic aperture sonar (SAS) images. The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient. The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed.

@@

Template matching is a common technique for classification of objects in synthetic aperture sonar (SAS) images. The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient. The challenge is to come up with a representative set of template images covering  the actual configurations of object and seabed. The use of a static, preset template library may be impractical and/or yield inaccurate results [Midelfart 4].

We have developed a sonar simulator that runs sufficiently fast to create templates adapted to the actual scene as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering. The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location. We assume rough, isotropic surfaces reflecting sound energy equally in all directions. This can be modeled with a Lambertian scattering model [1], where the intensity of the backscattered sound only depends on the incidence angle of the transmitted signal onto the model surface. When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.

Various simulators for high frequency, side-looking sonar imagery have been published, e.g. [Bell][Sammelmann]. Most implementations have however prioritized accurate acoustic modeling at the expense of execution speed. One exception is the SIGMAS+ simulator in [2,3] which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering. However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.

We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs). It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location. We assume rough, isotropic surfaces that reflects sound energy equally in all directions. This can be modeled with a Lambertian scattering model~\cite{Bell1997} where the intensity of the backscattered sound is said to depend only on the on incidence angle of the transmitted signal onto the model surface. When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixels to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.

By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}. They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model. However, while SIGMAS+ is geared more towards realistically looking sonar images, and include effects such as noise, we do not. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.

@@

\begin{itemize}
\item Midelfart/Midtgaard: Adaptive templates \cite{Midelfart2010}
\item Simulation of sidescan images (Bell thesis \cite{Bell1997}, pre-work by V.Blake \cite{Blake1993})
\end{itemize}

@@

Creating a 2D view of an arbitrary complex 3D scene is a non-trivial matter. This is why we decided to use the core OpenGL pipeline to do this for us. OpenGL is a popular, well-matured, and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics. It relieves us from the intricacies of handling vertices, faces and textures, as well as from projecting this onto a suitable 2D image plane. This section explains how we set up OpenGL to do this for us, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.

@@

Our models are stored in regular 3D model files. The simulator loads these with the Open Asset Import library (\texttt{assimp}). This is a portable open source library that supports loading a wide range of 3D model formats in a uniform manner. It handles the intricacies of the various model formats and outputs the model as a node tree. Each node contains data such as vertices, facets and textures, and since they are formatted in an OpenGL friendly way we simply iterate through the nodes and load it.

@@

To produce the sonar templates we assume a rough, isotropic surface that reflects energy equally in all directions. This permits us to use a Lambertian scattering model where the backscatter intensity depends only on the incidence angle~\cite{Zhang1999}. It does not consider observation angle or sound frequency\todo{is this unproblematic?}.

@@

\begin{table}[bh]
TODO: Probably not necessary to have this table...\\
\begin{tabular}[c]{l l l l l}\hline
\rowcolor{tabBlue} & cor   & ecor  & scor  & essum \\\hline
Standard template & 0.461 & 0.698 & 0.528 & 0.613 \\
Adaptive template & 0.673 & 0.818 & 0.807 & 0.813
\end{tabular}
\caption{Correlation scores. Description:\newline
cor - correlation between image and template calculated over all pixels in the image\newline
ecor - correlation calculated over the union of the echo pixels in the image\newline
scor - correlation calculated over the union of the echo pixels in image and template\newline
essum - 0.5*(ecor+scor). Average of the echo and shadows}
\end{table}

Reference correlation scores~\cite{Midelfart2010}

>>>>

\title{A GPU Sonar Simulator for Automatic Target Recognition}%
%
\author{Jo Inge Buskenes\firstAddress, Herman Midelfart\secondAddress, \O{}ivind Midtgaard\secondAddress}%
%
\begin{contact}
  \firstAddress Dept. of Informatics, P.O.Box 1080 Blindern, N-0316 OSLO\\
  \secondAddress Norwegian Defence Research Establishment (FFI) P.O.Box 25, NO-2027 Kjeller
\end{contact}%
%
\begin{contact}
Contact author: Jo Inge Buskenes\\
Dept. of Informatics, Univ. of Oslo, P.Box 1080 Blindern, N-0316 OSLO\\
\href{mailto:joibu@ifi.uio.no}{joibu@ifi.uio.no}
\end{contact}%
%
\begin{abstract}
Template matching is a common technique for automatic classification of objects in synthetic aperture sonar (SAS) images. The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient. The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed.

@@

Template matching is a common technique for classification of objects in synthetic aperture sonar (SAS) images. The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient. The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed\todo{objects? seabeds?}. The use of a static, predetermined template library may be impractical, yield inaccurate results, or both~\cite{Midelfart2010}.

We have developed a sonar simulator that runs sufficiently fast to create templates adapted to the actual scene as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering. The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location. We assume rough, isotropic surfaces reflecting sound energy equally in all directions. This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound only depends on the incidence angle of the transmitted signal onto the model surface. When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.

Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelmann2004}. Most implementations have however prioritized accurate acoustic modeling at the expense of execution speed. One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b} which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering. However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.

% We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs). It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location. We assume rough, isotropic surfaces that reflects sound energy equally in all directions. This can be modeled with a Lambertian scattering model~\cite{Bell1997} where the intensity of the backscattered sound is said to depend only on the on incidence angle of the transmitted signal onto the model surface. When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixels to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
% 
% By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}. They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model. However, while SIGMAS+ is geared more towards realistically looking sonar images, and include effects such as noise, we do not. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.

@@

@@

Creating a 2D view of an arbitrary complex 3D scene is a non-trivial matter. This is why we decided to use the core OpenGL pipeline to do this for us. OpenGL is a popular, well-matured, and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics. It relieves us from the intricacies of projecting vertices, faces and textures defined in a 3D space onto a suitable 2D image plane. This section explains how we set up OpenGL to do this for us, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.

@@

Our models are stored in regular 3D model files. The simulator loads these with the Open Asset Import library (\texttt{assimp}). This is a portable open source library that supports loading a wide range of 3D model formats in a uniform manner. Given a model file it outputs the model as a node tree where node contains data such as vertices, facets and textures, all formatted in an OpenGL friendly way.

@@

To produce the sonar templates we assume a rough, isotropic surface that reflects energy equally in all directions. This permits us to use a Lambertian scattering model where the backscatter intensity depends only on the incidence angle~\cite{Zhang1999}. It does not consider observation angle or sound frequency, but the purpose of creating templates this is needed.

@@
